---
title: "DecontamWorkflow_MiSeq382"
author: "Erin Dahl"
date: "2/13/2020"
output: html_document
---

## Set Workspace

```{r set workspace, message=FALSE}
# load libraries

library(phyloseq)
library(ggplot2)
library(decontam)
library(tidyverse)
library(patchwork)

## Load the dataset
load("mockDilutions.RData")

# Create profile of only expected sequences from the undiluted mock community sample

# Subset the undiluted mock microbial sample  (sample name 'D0')
mock_ps_pure <- subset_samples(mock_ps, sample_names(mock_ps)=='D0')

# Remove ASV's that are not present in the undiluted sample
mock_ps_pure <- prune_taxa(taxa_sums(mock_ps_pure)>0, mock_ps_pure)

# Change the SampleType and sample_names of the pure mock microbial community sample 

sample_data(mock_ps_pure)$SampleType <- 'MockCommunityProfile'
sample_names(mock_ps_pure) <- paste ('mc', sample_names(mock_ps_pure), sep = '_')

# display a summary of the new phyloseq object
mock_ps_pure
```

```{r}
# handle extra ASVs
mock_taxa<-as.matrix(mock_ps_pure@tax_table)
temp_taxaNames<-rownames(mock_taxa)

temp_taxaNamesMerged <- asv_key[asv_key$asv_name %in% temp_taxaNames,]

temp_taxa_seq <- temp_taxaNamesMerged$asv_sequence

adist(temp_taxa_seq)
```
For the [8,3] and [3,8] there is only 1 nt difference so this needs to be examined

```{r}

# identify number of sequences for each taxa
mock_counts<-as.data.frame(mock_ps_pure@otu_table)
# 

# need to re-label or keep



```

```{r}
# make a list of the top 9 abundant ASV taxa names (this is plausible for filtering since the 9 sequences we want to remove are present in low abundance)

mock_taxa = names(sort(taxa_sums(mock_ps_pure), decreasing = TRUE)[1:9])

# subset the taxa in mock_ps_pure so only the expected sequences are present

mock_ps_pure <- prune_taxa(mock_taxa, mock_ps_pure)
# removes unwanted OTUs

# create a phyloseq object that is normalized to 100 (relative abundance)
ps_norm <- transform_sample_counts(ps,function(x) 100* x/sum(x))
mock_ps_norm <- transform_sample_counts(mock_ps,function(x) 100* x/sum(x))

# Identify the proportion of each sample that is the expected mock community ASVs
ps_norm_exp <- prune_taxa(mock_taxa,ps_norm)

# Identify the proportion of each sample that is the expected mock community ASVs
ps_norm_exp <- prune_taxa(mock_taxa,ps_norm)

# Create a table with the dilution, number of reads per sample, and proportion of contaminants per sample
dilutionSummary <- data.frame(DilutionSeries = sample_names(ps),NumberOfReads = sample_sums(ps), PercentContaminants = 100-sample_sums(ps_norm_exp))

```

```{r }

# Original
#plot_bar(mock_ps_norm,fill='Genus', title = 'Mock Micribial Community Sequences') +theme(legend.position='none')

plot_bar(mock_ps_norm,fill='Genus', title = 'Mock Micribial Community Sequences')+theme(legend.position='none')

```
This figure is a lot different from the html file, not all bars reach 100% in the html plot. In the html the Qiagen subset of samples were graphed.

Since there is a lot of black in this, I am trying to figure out a way to remove this and have the bar graphs show shape a little bit more, might be coming from data noise, might not have corectly removed OTUs since the data is different

```{r}
# number of ASVs per sample

temp <- as.data.frame(ps@otu_table)
temp[temp > 0] <- 1
rowSums(temp)

# Evaluate the blank control

blank_ps_norm <- subset_samples(ps_norm,sample_names(ps_norm) %in% c('Blank'))
blank_ps_norm <- prune_taxa(taxa_sums(blank_ps_norm) > 0, blank_ps_norm)

#identify ASVs from mock community present in blank
table(taxa_names(blank_ps_norm) %in% mock_taxa)

blank_ps_norm %>% subset_taxa(taxa_names(blank_ps_norm) %in% taxa_names(mock_ps_pure)) %>% psmelt() %>% select(Abundance, Family, Genus)

# Collapse at the genus level, keeping unassigned genera

blank_ps_norm <- tax_glom(blank_ps_norm, "Genus", NArm = FALSE)

blank_ps_norm_melt <- psmelt(blank_ps_norm)
# Get genera present with >5% abundance
blank_ps_norm_melt %>% filter(Abundance > 5) %>% select(Abundance, Family, Genus)

# Get genera present with abundance between 1 and 5%
blank_ps_norm_melt %>% filter(Abundance <= 5) %>% filter(Abundance > 1) %>% select(Abundance, Family, Genus)
```

To evaluate the contaminant ASVs, we create a phyloseq object only containing the contaminant ASVs.

```{r}
# create a list of unexpected sequences (contaminants)

# create a list of all ASV taxa names
contaminant_taxa<-taxa_names(mock_ps)
# remove the expected mock community ASV taxa names
contaminant_taxa <- contaminant_taxa[!(contaminant_taxa %in% mock_taxa)]

# create a phyloseq object that only contains the contaminant sequences (for use with sourcetracker)
contaminants_ps<-prune_taxa(contaminant_taxa,mock_ps)
contaminants_ps<- prune_taxa(taxa_sums(contaminants_ps)>0,contaminants_ps)

# change the sample names to indicate that these samples only contain contmaminant ASVs
sample_names(contaminants_ps)<-paste('con',sample_names(contaminants_ps),sep = '_')
sample_data(contaminants_ps)$SampleType<-'ContaminantProfile'

# create phyloseq object from normalized data to summarize contamiant contribution
contaminants_ps_norm<-prune_taxa(contaminant_taxa,mock_ps_norm)
contaminants_ps_norm<- prune_taxa(taxa_sums(contaminants_ps_norm)>0,contaminants_ps_norm)

```

Now, we use the contaminants_ps object to evaluate the number contaminant ASVs that are present in the blank control sample.

```{r}
# Number of contaminant ASVs across dilution series
print(paste('Total number of contaminant ASVs', length(taxa_names(contaminants_ps))))

# Abundance of contaminant genera diltuion series
contaminant_genera <- tax_glom(contaminants_ps_norm, 'Genus', NArm = FALSE) %>% 
  psmelt() 

# Maximum abundance of contaminant genera per diltuion sample
contaminant_genera %>%
  group_by(Sample) %>%
  filter(Abundance == max(Abundance)) %>% 
  select(Sample, Abundance)
```

```{r}
# Create list of contamiant ASVs with abundance > 1% (Supplemental Table 2)
contam_table <-contaminants_ps_norm %>% 
  psmelt() %>%
  group_by(OTU) %>%
  mutate(max_abundance = max(Abundance)) %>% 
  filter(max_abundance > 1) %>%
  mutate(in_n_samples = sum(Abundance > 0)) %>%
  mutate(in_blank = OTU %in% taxa_names(blank_ps_norm)) %>%
  mutate(asv_sequence = asv_key[asv_key$asv_name == OTU,]$asv_sequence) %>%
  select( Phylum, Class, Order, Family, Genus,max_abundance, in_n_samples, in_blank, asv_sequence) %>% 
  unique()
```
```{r}
# number of ASVs present in blank sample
print(paste('Number of contaminant ASVs also present in blank', length(intersect(taxa_names(contaminants_ps),taxa_names(blank_ps)))))
```

```{r}
# create a list of contaminants taxa that are not present in the blank control
contaminant_taxa_no_blank<-taxa_names(contaminants_ps)
contaminant_taxa_no_blank <- contaminant_taxa_no_blank[!(contaminant_taxa_no_blank %in% taxa_names(blank_ps))]

# Create  a binary list of contaminant ASVs indicating if the ASV is present in the blank control (1) or not (0)
contaminants_in_blank <- data.frame(matrix(1, ncol = length(taxa_names(contaminants_ps)), nrow = 1))
colnames(contaminants_in_blank) <- taxa_names(contaminants_ps)
contaminants_in_blank[,contaminant_taxa_no_blank] <- 0 
contaminants_in_blank <- t(contaminants_in_blank)
```

We found that there are a lot of contaminant ASVs present in the dilution series that aren’t in the blank control sample.

We also determine the proportion of contaminant ASVs that are not in found in the blank control to help us identify the actual impact of these.

```{r}
# Identify the contribution per sample of contaminants that are not present in blanks
# generate a phyloseq object with contaminants only normalized to 100 
contaminant_ps_norm <- transform_sample_counts(contaminants_ps,function(x) 100* x/sum(x))
contaminant_no_blanks<-prune_taxa(contaminant_taxa_no_blank,contaminant_ps_norm)
```

```{r}

# Original
# plot_bar(contaminant_no_blanks,fill='Genus',title = 'Contamiants No Blank ASVs') + theme(legend.position='none')

plot_bar(contaminant_no_blanks,fill='Genus', title = 'Contamiants No Blank ASVs') + geom_bar(stat="identity") + ylim(c(0,100)) + theme(legend.position='none')

```

```{r define functions}
isContmFreqThr<- function(physeq,thr){
    contam.freq <- isContaminant(physeq, method="frequency", conc="DNA_conc",threshold=thr)
    print(table(contam.freq$contaminant))
    physeq.nc.freq  <- prune_taxa(!contam.freq$contaminant, physeq)
    # Test performance on Mock Community samples
    mock_ps.nc.freq<-subset_samples(physeq.nc.freq,SampleType== 'MockCommunity')
    mock_ps.nc.freq<-prune_taxa(taxa_sums(mock_ps.nc.freq)>0,mock_ps.nc.freq)
    # Normalize to 100
    mock_ps.nc.freq<-transform_sample_counts(mock_ps.nc.freq,function(x) 100* x/sum(x))
    # Remove non-mock sequences to evaluate how accurately the 'true' sequences were recovered
    mock_ps.nc.freq<-prune_taxa(mock_taxa,mock_ps.nc.freq)
    print(100 - sample_sums(mock_ps.nc.freq)) 
    plot_bar(mock_ps.nc.freq,fill='Genus',title=paste('Frequency, threshold = ',thr)) +theme(legend.position='none') + scale_fill_brewer(palette="Paired")
    }
# Plot expected sequences - need to pass phyloseq object with only the desired subject
# use subject_ps <- subset_samples(physeq.nc,SampleID == "value of SampleID")
expSeq<- function(physeq.nc,subjectID){
    # define taxa that should be present based on Neat sample
    neat_ps.nc<-subset_samples(physeq.nc,SampleDescription == 'MC_Neat')
    neat_ps.nc<-prune_taxa(taxa_sums(neat_ps.nc)>0,neat_ps.nc)
    neat_taxa <-rownames(neat_ps.nc@tax_table)
    # Normalize to 100 
    physeq.nc<-transform_sample_counts(physeq.nc,function(x) 100* x/sum(x))
    # Remove non-neat sequences to evaluate how accurately the true sequences were recovered
    physeq.nc<-prune_taxa(neat_taxa,physeq.nc)
    print(100 - sample_sums(physeq.nc)) 
    plot_bar(physeq.nc,fill='Genus',title=paste('Expected Sequences for ',subjectID)) +theme(legend.position='none') + scale_fill_brewer(palette="Paired")
    }

isContmPrevThr<- function(physeq,thr){
    contam.prev <- isContaminant(physeq, method="prevalence", neg = 'is.neg',threshold=thr)
    print(table(contam.prev$contaminant))
    physeq.nc.prev  <- prune_taxa(!contam.prev$contaminant, physeq)
    # Test performance on Mock Community samples
    mock_ps.nc.prev<-subset_samples(physeq.nc.prev,SampleType== 'MockCommunity')
    mock_ps.nc.prev<-prune_taxa(taxa_sums(mock_ps.nc.prev)>0,mock_ps.nc.prev)
  # Normalize to 100
    mock_ps.nc.prev<-transform_sample_counts(mock_ps.nc.prev,function(x) 100* x/sum(x))
  # Remove non-mock sequences to evaluate how accurately the 'true' sequences were recovered
    mock_ps.nc.prev<-prune_taxa(mock_taxa,mock_ps.nc.prev)
    print(100 - sample_sums(mock_ps.nc.prev)) 
  quartz()
    plot_bar(mock_ps.nc.prev,fill='Genus',title=paste('prevalence, threshold = ',thr)) +theme(legend.position='none') + scale_fill_brewer(palette="Paired")    }

isContmComboThr<- function(physeq,thr){
    contam.combo <- isContaminant(physeq, method="combined", neg="is.neg", conc="DNA_conc",threshold = thr)
    print(table(contam.combo$contaminant))
    physeq.nc.combo  <- prune_taxa(!contam.combo$contaminant, physeq)
    # Test performance on Mock Community samples
    mock_ps.nc.combo<-subset_samples(physeq.nc.combo,SampleType== 'MockCommunity')
    mock_ps.nc.combo<-prune_taxa(taxa_sums(mock_ps.nc.combo)>0,mock_ps.nc.combo)
  # Normalize to 100
    mock_ps.nc.combo<-transform_sample_counts(mock_ps.nc.combo,function(x) 100* x/sum(x))
  # Remove non-mock sequences to evaluate how accurately the 'true' sequences were recovered
    mock_ps.nc.combo<-prune_taxa(mock_taxa,mock_ps.nc.combo)
    print(100 - sample_sums(mock_ps.nc.combo))
    plot_bar(mock_ps.nc.combo,fill='Genus',title=paste('Combination, threshold = ',thr)) +theme(legend.position='none') + scale_fill_brewer(palette="Paired")    }

```


```{r}
ps_norm <- transform_sample_counts(ps,function(x) 100* x/sum(x))
mock_ps_norm <- transform_sample_counts(mock_ps,function(x) 100* x/sum(x))

# Identify the proportion of each sample that is the expected mock community ASVs
ps_norm_exp <- prune_taxa(mock_taxa,ps_norm)

# Create a table with the dilution, number of reads per sample, and proportion of contaminants per sample
dilutionSummary <- data.frame(DilutionSeries = sample_names(ps),NumberOfReads = sample_sums(ps), PercentContaminants = 100-sample_sums(ps_norm_exp))

# Create a variable to indicate the sample order of the plots
dilutions<-c('D0','D1','D2','D3','D4','D5','D6','D7','D8', 'Blank')

# Create plots to summarize these data
## Plot Figure 1A - number of reads per sample across dilution series
dilutionSummary %>% subset(DilutionSeries %in% dilutions[1:9]) %>%
  ggplot(.,aes(x=DilutionSeries, y= NumberOfReads)) + geom_bar (stat="identity", fill="steelblue") +
  theme_minimal() +scale_x_discrete(limits = dilutions) +
  labs (title = "Number of Reads per Dilution", x= "Dilution Series", y= "Number of Reads") +
  theme (axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
```

```{r Percent Contaminants Graph}
## Plot Figure 1B - Percent of contaminants across dilution series
dilutionSummary %>% subset(DilutionSeries %in% dilutions[1:9] ) %>%
  ggplot(., aes(x = DilutionSeries, y = PercentContaminants)) + geom_point(size = 3) +
  scale_x_discrete(limits = dilutions) +   
  labs(x = "Dilution Series", y = "Percent Contaminants") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
```
Results on Mock communtiy dilution with different thresholds

```{r}

# Apply frequency method for identifying contaminants at different Thresholds
plotfreq.1 <- isContmFreqThr(mock_ps, 0.1) + theme(plot.title= element_text(size = 10))
plotfreq.2 <- isContmFreqThr(mock_ps, 0.2) + theme(plot.title= element_text(size = 10))
plotfreq.3 <- isContmFreqThr(mock_ps, 0.3) + theme(plot.title= element_text(size = 10))
plotfreq.4 <- isContmFreqThr(mock_ps, 0.4) + theme(plot.title= element_text(size = 10))
plotfreq.5 <- isContmFreqThr(mock_ps, 0.5) + theme(plot.title= element_text(size = 10))
```

```{r}
#(plotfreq.1 + plotfreq.2 +plotfreq.3)/ (plotfreq.4 + plotfreq.5)

plotfreq.1 + plotfreq.2 +plotfreq.3 + plotfreq.4 + plotfreq.5
```

```{r}
contam.freq <- isContaminant(ps, method="frequency", threshold=0.5, conc="DNA_conc")

g_norm <- tax_glom(ps_norm,'Genus')

ps.noncontam.freq <- prune_taxa(!contam.freq$contaminant, ps)
g.noncontam.freq <- tax_glom(ps.noncontam.freq,'Genus',NArm = FALSE)

# plot results for each mock and UTI sample diltion series
ps_D0 <-subset_samples(g_norm, X.SampleID == 'D0')
ps_D1 <-subset_samples(g_norm, X.SampleID == 'D1')
ps_D2 <-subset_samples(g_norm, X.SampleID == 'D2')

ps_D0.nc <-subset_samples(g.noncontam.freq, X.SampleID == 'D0')
ps_D1.nc <-subset_samples(g.noncontam.freq, X.SampleID == 'D1')
ps_D2.nc <-subset_samples(g.noncontam.freq, X.SampleID == 'D2')
#ps_mock.nc <- subset_samples(g.noncontam.freq, SampleID == 'MicrobialCommunity_Qiagen')

expSeq(ps_D0,'D0 original')

```
