---
title: "DecontamWorkflow_Miseq401"
author: "Erin Dahl"
date: "2/20/2020"
output: html_document
---

MiSeq401

```{r}
#load libraries
library(phyloseq)
library(dplyr)
library(ggplot2)
library(here)
library(decontam)
library(stringdist)
library(RColorBrewer)
library(tidyverse)
library(stringdist)
```

Load data through path
```{r}
# load path
ps_path <- here('mockDilutions.RData')

if(file.exists(ps_path)){
  load(ps_path) 
} else {
  stop(paste('Cannot find data, please check path ',ps_path )) 
}

```


# Summarize/Review data

``` {r}
ps

samColNames <- ps@sam_data@names

# Checks to make sure col headers for SampleType/Origin is consistent and changes it if needed to Sample Origin

# string function to make sample name all lower case
# error message indicating 
# reassign var names

if (!('Sample_Origin' %in% samColNames)) {
  print(paste("Sample Origin from sam data not identified, please assign to temp char variable __ "))
  
  sampleOriginName = "SampleType"
  #Renames SampleType to Sample_Origin
  #samColNames <- str_replace(samColNames, "SampleType", "Sample_Origin")
  #ps@sam_data@names <- samColNames
}

# SampleSums is the number of reads per sample, add this data to sample_data
sample_data(ps)$SampleSums <- sample_sums(ps)

# review sample types in dataset 
table(sample_data(ps)$SampleType)
```

A faceted plot of number of reads per sample per sample type/ sample origin

```{r}
# ggplot(sample_data(ps), aes(as.factor(SampleType), SampleSums)) + geom_boxplot()
ggplot(sample_data(ps), aes(as.factor(sampleOriginName), SampleSums)) + geom_boxplot(aes(fill= as.name("sampleOriginName"))) + xlab("Sample Origin") + ylab("Sample Sums") 

```

```{r}
ggplot(sample_data(ps), aes(as.factor(SampleType), SampleSums)) + geom_boxplot(aes(fill=SampleType),) + xlab("Sample Type") + ylab("Sample Sums") + geom_dotplot(binaxis='y', stackdir='center',
                 position=position_dodge(1))

# add binwidth arg to geom_dotplot to get rid of warning. It will make the dots much smaller
```

A boxplot (or similar) of number of  reads per sample per sample type/ sample origin

```{r}
# Evaluate the number of reads per sample
ggplot(sample_data(ps), aes(x=SampleSums, color=SampleType)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")  +
  ggtitle('Number of Reads per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5)) + xlab("Sample sums") + ylab("Count")
```

```{r}
ggplot(sample_data(ps), aes(x=SampleSums, color=SampleType)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")  +
  facet_wrap("SampleType") +
  ggtitle('Number of Reads per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5)) + xlab("Sample sums") + ylab("Count")
```

A faceted plot of number of the DNA concentration per sample type/ sample origin
```{r}
ggplot(sample_data(ps), aes(as.factor(SampleType), DNA_conc)) + geom_boxplot(aes(fill=SampleType)) + xlab("Sample Type") + ylab("DNA concentration") 

```

A boxplot (or similar) of of the DNA concentration per sample type/ sample origin

```{r}
ggplot(sample_data(ps), aes(x=DNA_conc, color=SampleType)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")  +
  ggtitle('DNA concentration per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
ggplot(sample_data(ps), aes(x=DNA_conc, color=SampleType)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")  +
  facet_wrap("SampleType") +
  ggtitle('DNA concentration per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5))
```

Number of reads by DNA concentration per sample by sample type

```{r}
ggplot(sample_data(ps), aes(x=DNA_conc, y = SampleSums, color=SampleType)) +
  geom_point() + ggtitle('Number of reads by DNA concentration per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
ggplot(sample_data(ps), aes(x=DNA_conc, y = SampleSums, color=SampleType)) +
  geom_point() +
  facet_wrap("SampleType") +
  ggtitle('Number of reads by DNA concentration per sample by sample type') +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
# Subset the undiluted mock microbial sample  (sample name 'D0')
mock_ps_pure <- subset_samples(mock_ps, sample_names(mock_ps)=='D0')

# Remove ASV's that are not present in the undiluted sample
mock_ps_pure <- prune_taxa(taxa_sums(mock_ps_pure)>0, mock_ps_pure)

mock_taxa <- taxa_names(mock_ps_pure)

# identify if any expected ASVs are split into two (just one or two nt off)

mock_taxa<-rownames(as.matrix(mock_ps_pure@tax_table))

temp_taxaNamesMerged <- asv_key[asv_key$asv_name %in% mock_taxa,]

temp_taxa_seq <- temp_taxaNamesMerged$asv_sequence

adist(temp_taxa_seq)


```


```{r}
# Normalize the mock microbial sequences to 100 to identify the % of contaminants in the dilution series
mock_ps_norm <- transform_sample_counts(mock_ps,function(x) 100* x/sum(x))

# Plot the expected mock microbial communtiy sequences from the mock microbial community dilution series
mock_ps_norm_exp <- prune_taxa(mock_taxa,mock_ps_norm)
plot_bar(mock_ps_norm_exp,fill = 'Genus') + theme(legend.position='none')
```



```{r}
# Normalize the full dataset to relative abunance
ps_norm <- transform_sample_counts(ps,function(x) 100* x/sum(x))

```

Define functions for evaluating decontam
```{r}
 expCompBarPlot <- function(physeq, exp_taxa, title){
   ## physeq - phyloseq object that will be plotted
   ## exp_taxa - taxa that are expected to be in the mock community 
   ## title - title for plot
   
#set up data_table
  data_table <- as.data.frame(t(physeq@otu_table))

  data_table$reference = FALSE
  data_table$reference[rownames(data_table) %in% exp_taxa] = TRUE
  sample_names <- sample_names(physeq)
  data_table$id <- paste0('ASV_', 1:nrow(data_table))
  dilution_labels <- sample_data(physeq)$Dilutions


# define the colors to use for reference and non-reference OTUs/ASVs
  set.seed(444)
  ref_colors <- brewer.pal(sum(data_table$reference), "Paired")
  other_colors <- sample(grey.colors(5, start = 0.5, end = 0.9), sum(!data_table$reference), replace = TRUE)

# add a color variable to the data table
  data_table$color <- rep(NA, nrow(data_table))
  data_table$color[data_table$reference] <- ref_colors
  data_table$color[!data_table$reference] <- other_colors

# reshape the data table into a ggplot-friendly format, by gathering samples into a single column called "count"

  color_gg <- data_table %>% select(id, sample_names, color) %>% gather("sample", "count", sample_names)
  legend_color <- c(bright = ref_colors[2], dull = other_colors[2])
  data_gg <- data_table %>% gather("sample", "count", sample_names)

  data_gg <- inner_join(data_gg,color_gg)

# create the composition bar plot
  comp_bar <- ggplot(data_gg, aes(x = sample, y = count)) +
    geom_col(aes(fill = color, group = reference, alpha = ifelse(reference, "bright", "dull")), width = 0.7, position = position_fill()) +
    scale_fill_identity(guide = FALSE) +
    scale_alpha_manual(name = "Sequence type",
                     labels = c("expected sequences", "other"),
                     values = c(bright = 1, dull = 1),
                     guide = guide_legend(override.aes = list(fill = c(ref_colors[4], "#AEAEAE")),
                                          keywidth = NULL, keyheight = NULL)) +
    labs(title = title, x = "sample", y = "Relative Abundance") +
    theme(legend.position = "right", legend.title = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
  comp_bar
}
```

## Decontam - frequency method

Test levels of decontam to remove contaminants

```{r}
decontamTestFreq <- function(physeq, exp_taxa, thr){
  contam.freq <- isContaminant(physeq, method="frequency", threshold=thr, conc="DNA_conc")
  print(table(contam.freq$contaminant))
  ps.noncontam.freq <- prune_taxa(!contam.freq$contaminant, physeq)
  mock_test <- subset_samples(ps.noncontam.freq,SampleType== 'MockCommunity') ##check if this function change is valid for the mock community
  expCompBarPlot(mock_test, exp_taxa,paste('Mock community decontam results \n frequency method threshold = ',thr))

}

decontamTestFreq(ps,mock_taxa, 0)
```

```{r}

decontamTestFreq(ps,mock_taxa, 0.1)
```

##Function for comparing frequency Thresholds with plot of ASV's removed

```{r}
#Function Erin Wrote for comparing frequency thresholds with plot of ASV's removed

decontamFreqThrCounts <- function(physeq){
  
  freqVector = c()
  thrVector <- seq(.1, .9, by=0.1)
  
  for ( i in 1:9)
  {
    contam.freq <- isContaminant(physeq, method="frequency", threshold=thrVector[i], conc="DNA_conc") 
    #print(table(contam.freq$contaminant))
    #print(sum(contam.freq$contaminant == TRUE))
    freqVector[i] <- sum(contam.freq$contaminant == TRUE)
  }
  
  asvFreqSums <- data.frame (freqVector, thrVector)
  
  
  ggplot(asvFreqSums, aes(x = thrVector, y = freqVector))+ geom_bar(stat="identity", fill="steelblue", position = position_dodge() ) + theme_minimal() + labs (x = "Decontam Threshold", y = "Number ASVs identified as contaminants") + scale_x_discrete(limits = thrVector) 
 
}

decontamFreqThrCounts(ps)
```

##Function for comparing frequency Thresholds with plot of ASV's removed

```{r}
#Function Erin Wrote for comparing frequency thresholds with plot of percent ASV's removed

decontamFreqThrPercents <- function(physeq){
  
  freqVector = c()
  thrVector <- seq(.1, .9, by=0.1)
  totalTaxa <- ntaxa(physeq)
  
  for ( i in 1:9)
  {
    contam.freq <- isContaminant(physeq, method="frequency", threshold=thrVector[i], conc="DNA_conc") 
    #print(table(contam.freq$contaminant))
    #print(sum(contam.freq$contaminant == TRUE))
    freqVector[i] <- (sum(contam.freq$contaminant == TRUE) / totalTaxa)
  }
  
  asvFreqSums <- data.frame (freqVector, thrVector)
  
  
  ggplot(asvFreqSums, aes(x = thrVector, y = freqVector))+ geom_bar(stat="identity", fill="steelblue", position = position_dodge() ) + theme_minimal() + labs (x = "Decontam Threshold", y = "Percent ASVs identified as contaminants") + scale_x_discrete(limits = thrVector) #+ scale_y_continuous(limits=c(0, 100))
  
  #add limit y axis 100
 
}

decontamFreqThrPercents(ps)
```


